---
title: "Profiling: reduce, cluster, characterise"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling: reduce, cluster, characterise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
## clustering needs the cluster package, which illumex suggests
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5,
                      eval = requireNamespace("cluster", quietly = TRUE))
```

Three steps that are usually taken together: find the few directions a set of
correlated columns shares, group the rows in that space, and say what
distinguishes each group.

```{r}
library(illumex)
```

The examples use simulated survey data: 300 people, with their age, income,
household size, number of visits and work, drawn so that younger people on
lower incomes, people on higher incomes in larger households, and older
people, mostly retired, who visit more often, are all in it -- though
nothing in the data says who is which. To keep this page quick to build,
the clustering uses 25 resamples (`B`), where the default is 100.

Two dimensions are kept (`ndim = 2`). With five columns, the dimensions past
the first two mostly carry the levels of `work` on their own, and the gap
statistic then splits the groups along them: kept at five, it chose 8
clusters here. Choose `ndim` from the scree plot below, not by default.

```{r}
set.seed(1)
n <- 300
g <- sample(1:3, n, TRUE, prob = c(0.4, 0.35, 0.25))
d <- data.frame(
  age = round(c(31, 47, 71)[g] + rnorm(n, 0, 6)),
  income = round(c(38, 72, 31)[g] + rnorm(n, 0, 9), 1),
  household = 1 + rpois(n, c(1.2, 2.4, 0.4)[g]),
  visits = rpois(n, c(2, 3, 6)[g]),
  work = factor(ifelse(runif(n) < c(0.9, 0.95, 0.15)[g], "employed",
                       ifelse(g == 3, "retired", "studying"))))
```

## The whole thing in one call

```{r}
pr <- ilm_profile(d, ndim = 2, B = 25, seed = 1)
pr
ilm_plot_profile(pr)
```

`ilm_profile()` runs `ilm_reduce()`, then `ilm_cluster()` on the dimensions it
produced, then characterises the clusters. The three are also available
separately, which is what you want as soon as you disagree with one of its
choices.

## Reducing

```{r}
rr <- ilm_reduce(d, ndim = 2)
rr
ilm_plot_reduce(rr)
ilm_plot_reduce_scree(rr)
ilm_plot_reduce_contrib(rr)
```

The method follows the column types without being told: PCA for all-numeric
data, multiple correspondence analysis for all-categorical, and factor analysis
of mixed data when both are present. All three are the same decomposition with
different scaling, which is why one function covers them.

The number of dimensions available is `min(n - 1, p)` for PCA and the mixed
case, and the number of levels less the number of variables for MCA. Asking for
more than exist is an error rather than a smaller answer, which is worth knowing
because an earlier version computed that ceiling as `p - 1` and failed outright
on any two-column selection.

### A different loss per column

FAMD reaches mixed data by one-hot encoding the categories, scaling the
indicators and running PCA on the result. That is fast, has a closed form, and
treats a category as though it were Gaussian -- a reconstruction can come back
at `-0.3`, which is not a category.

```{r}
rr <- ilm_reduce(d, ndim = 2, method = "glrm", lambda = 2)  # lambda given: no search
```

A generalised low rank model keeps the low-rank structure and changes the loss:
quadratic for a numeric column, logistic for a binary one, multinomial across
the levels of a categorical one, Poisson for a count. Each column is then
reconstructed on its own scale, and a category comes back as a category.

It costs an iterative fit where FAMD has a decomposition, so it is an option
rather than the default. On all-numeric data the two are the *same model* --
with quadratic loss everywhere and no penalty a GLRM is principal components,
verified against the SVD to 1e-16 -- so there is nothing to gain there. What it
adds is mixed data, and `illume::ilm_impute(method = "glrm")` uses the same machinery to
impute categorical columns.

See `?ilm_glrm` for the ridge penalty, which is chosen by cross-validation
because a fixed value is wrong across sizes and missingness rates.

## Clustering

```{r}
cl <- ilm_cluster(rr, B = 25, seed = 1)
cl
ilm_plot_cluster(cl)
ilm_plot_cluster_gap(cl)
```

`k` is chosen by the gap statistic rather than by eye. The result also reports
what a cluster solution usually hides: how stable each cluster is under
resampling (bootstrap Jaccard), how many rows sit ambiguously between two of
them (silhouette width), and whether any cluster is too small to be a finding.

A cluster that appears in 45% of bootstrap resamples is not a group, and the
output says so rather than leaving the reader to assume that everything printed
is solid.

## Characterising

The step people usually do by hand, badly. Each cluster is compared with all
rows on every variable the clustering used, and described in that variable's
own units: a number by the middle 50% of its values against everyone's, a
category by the value whose share in the cluster sits furthest from its share
among all rows, a date by the middle 50% of its values and by any stretch of
the week, month or year where the cluster stands out.

```{r}
cat(pr$summary, sep = "\n\n")   # a paragraph per cluster
pr$characterization               # every comparison behind them
pr$frequencies                    # every value of every category, per cluster
pr$by_cluster                     # ilm_describe_all(), by cluster
```

The variables are ranked by a v-test -- how far the cluster's mean or share sits
from everyone's, in standard errors of a subset that size -- and named, strongest
first, when it clears 1.96 and the difference is big enough to matter: 0.2
standard deviations for a number, 10 percentage points for a share. With many
rows a chance difference clears 1.96 on its own, which is what the second bar is
for. Variables that set no cluster apart are said to, after the paragraphs.

That is what a cluster is for: "older, mostly retired, living alone or in twos"
is a finding, and "cluster 3" is not. The v-test ranks the differences; it does
not test them, since the clusters were found from these same variables.

## What this is, and is not

This is a lighter-weight reimplementation of the `FactoMineR::HCPC()` plus
`catdes()` workflow, with the stability reporting that pipeline does not do, and
without the dependency. It is exploratory. The groups it finds are descriptions
of the sample rather than populations, and treating a cluster label as a
variable in a subsequent model -- then testing whether the clusters differ on
the variables that defined them -- is circular.

If the question is whether a *known* grouping differs, that is a model, and `illume`'s:

```{r, eval = FALSE}
illume::ilm_model(y ~ group, data = d, family = "gaussian")
```

## Structure in the missingness

The same three steps run on the missingness pattern rather than the values.
In R's `airquality`, two columns have missing values:

```{r}
ilm_reduce_na(airquality)
pr_na <- ilm_profile_na(airquality, k_max = 4, B = 25, seed = 1)
pr_na
ilm_plot_profile_na(pr_na)
```

A questionnaire where income and savings go missing together, and separately
from the health block, has structure in its non-response. Knowing that before
choosing an imputation strategy is worth more than most of what the imputation
then does.

## See also

`?ilm_reduce`, `?ilm_cluster`, `?ilm_profile`, `?ilm_glrm`,
`vignette("anomaly-detection")` for rows that fit no group, and
`vignette("missing-data", package = "illume")`.
