---
title: "Validated against FactoMineR and PCAmixdata"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(illumex)
## differences this small are what the page is about, so they are written
## out rather than rounded to zero
sci <- function(x) formatC(x, format = "e", digits = 1)
## the stored references live with the tests in the package's source
fixtures <- file.path("..", "..", "tests", "testthat", "fixtures")
source(file.path("..", "..", "tests", "testthat", "helper-famd-cases.R"))
```

`ilm_reduce()`, and so `ilm_cluster()` and `ilm_profile()`, compute their
dimensions by factor analysis of mixed data (Pagès 2004): principal
components when every column is a number, multiple correspondence analysis
when every column is a category, and the mixed method in between. illumex
computes it itself. It used to call `PCAmixdata::PCAmix()`, and before
relying on its own it was held to that and to `FactoMineR::FAMD()`, the two
established implementations. The outputs of both are stored with the
package's tests, so every check of the package repeats the comparison
without either being installed. This page runs it again as the site is
built.

## Against PCAmixdata

Four kinds of data from `ilm_sim()`: mixed (four numbers, two categories),
numbers only, categories only, and mixed with missing values. The largest
difference from what `PCAmix()` gave, on every eigenvalue, every coordinate
of every row on every retained dimension, and every squared loading:

```{r pcamix, echo = FALSE}
ref <- utils::read.csv(file.path(fixtures, "pcamix_cases.csv"), comment.char = "#",
                       colClasses = c(row = "character"))
cases <- famd_cases()
tab <- do.call(rbind, lapply(names(cases), function(nm) {
  cs <- cases[[nm]]; r <- ref[ref$case == nm, ]
  f <- illumex:::ilm_famd(cs$quanti, cs$quali, ndim = cs$ndim)
  eig <- r$value[r$what == "eigenvalue"]
  co <- r[r$what == "coord", ]
  d_co <- max(vapply(unique(co$dim), function(k) {
    theirs <- co$value[co$dim == k]; mine <- f$ind$coord[, k]
    max(abs(sign(sum(mine * theirs)) * mine - theirs))
  }, 0))
  sq <- r[r$what == "sqload", ]
  d_sq <- max(abs(mapply(function(v, k) f$sqload[v, k], sq$row, sq$dim) - sq$value))
  data.frame(data = nm, rows = nrow(f$ind$coord),
             eigenvalues = sci(max(abs(f$eig[seq_along(eig), 1] - eig))),
             coordinates = sci(d_co), loadings = sci(d_sq))
}))
knitr::kable(tab, align = "lrrrr")
```

A coordinate's sign is arbitrary in any implementation -- the same
dimension read from the other end -- so coordinates are compared after
turning each of illumex's dimensions to face the same way as PCAmix's.
illumex fixes its own signs by a rule, so the same data always give the
same coordinates: the column that loads most on a dimension loads
positively.

## Against FactoMineR

Five columns of `ilm_sim()`, through `ilm_reduce()` itself, against what
`FactoMineR::FAMD()` gave: its first five eigenvalues, and every row's
coordinates on the first three dimensions.

```{r factominer, echo = FALSE}
fa <- utils::read.csv(file.path(fixtures, "famd_ilm_sim.csv"))
d <- ilm_sim()[c("score", "income", "visits", "claims", "grp")]
r <- suppressMessages(ilm_reduce(d, ndim = 5))
co <- vapply(1:3, function(k) {
  theirs <- fa$value[fa$what == "coord" & fa$dim == k]
  mine <- r$ind_coord[[paste0("dim", k)]]
  max(abs(sign(sum(mine * theirs)) * mine - theirs))
}, 0)
knitr::kable(data.frame(
  quantity = c("eigenvalues 1 to 5", "coordinates, dimensions 1 to 3"),
  largest_difference = sci(c(max(abs(r$eig$eigenvalue[1:5] - fa$value[fa$what == "eigenvalue"])),
                             max(co)))), align = "lr")
```

`ilm_profile()` describes each cluster by v-tests, as `FactoMineR::catdes()`
does: for a number, how far the cluster's mean sits from the overall mean in
standard errors; for a category, the hypergeometric probability of the
cluster's count on the normal scale. Against `catdes()` on a fixed partition
of the same data:

```{r catdes, echo = FALSE}
cd <- utils::read.csv(file.path(fixtures, "catdes_ilm_sim.csv"), stringsAsFactors = FALSE)
dd <- ilm_sim()
cl <- cut(dd$score, stats::quantile(dd$score, 0:3 / 3), include.lowest = TRUE, labels = FALSE)
num <- cd[is.na(cd$level), ]; cat_ <- cd[!is.na(cd$level), ]
v_num <- mapply(function(k, v) illumex:::ilm_vtest_mean(dd[[v]], cl == k), num$cluster, num$variable)
v_cat <- mapply(function(k, v, l) illumex:::ilm_vtest_share(dd[[v]] == l, cl == k),
                cat_$cluster, cat_$variable, cat_$level)
knitr::kable(data.frame(
  v_test = c("numeric variables", "categories"),
  compared = c(nrow(num), nrow(cat_)),
  largest_difference = sci(c(max(abs(v_num - num$v_test)), max(abs(v_cat - cat_$v_test))))),
  align = "lrr")
```

## Speed

Measured once by `dev/studies/famd_own.R` in the package's source, on one
core, as the median ratio of illumex's time to PCAmix's over repeated runs
(below 1 is faster); FactoMineR's is beside it:

| rows | columns | illumex / PCAmix | FactoMineR / PCAmix |
|---:|---:|---:|---:|
| 200 | 12 | 1.25 | 1.50 |
| 2,000 | 12 | 0.35 | 0.83 |
| 2,000 | 120 | 0.62 | 0.97 |
| 20,000 | 12 | 0.19 | 0.80 |
| 20,000 | 120 | 0.51 | 0.96 |
| 100,000 | 20 | 0.22 | 0.81 |

## Checking it yourself

The comparisons above run in the package's tests, `test-famd.R` and
`test-factominer-agreement.R`, on every check. The references were recorded
by `dev/studies/make_pcamix_fixtures.R` (PCAmixdata 3.1) and
`dev/studies/make_factominer_fixtures.R` (FactoMineR 2.11); run them again
if either package changes its method.

## References

Pagès, J. (2004). Analyse factorielle de données mixtes. *Revue de
Statistique Appliquée* 52(4), 93-111.

Chavent, M., Kuentz-Simonet, V., Labenne, A. and Saracco, J. (2014).
Multivariate analysis of mixed data: the PCAmixdata R package.
arXiv:1411.4911.

Lê, S., Josse, J. and Husson, F. (2008). FactoMineR: an R package for
multivariate analysis. *Journal of Statistical Software* 25(1), 1-18.
